---
title: "todd_5530_m8"
author: "Reagan Todd"
format:
  html:
    embed-resources: true
execute: 
  error: true
---
### This is a demonstration of how to join tables in SQL and select rows.                             Created 2026-04-03 by Reagan Todd.                                                                       The copyright is for educational uses only. 

```{r}
#| label: setup
#| message: false
#| warning: false
R.version.string
Sys.Date()
library(tidyverse)
library(DBI)
library(RSQLite)
library(readr)
```

# Use the MIMIC-III Schema.
# Q1. Retrieve patient gender and admission time for 10 rows.                                           Hint: Join patients and admissions to get info.
## First we need to ensure the database is loaded correctly.
```{r}
#| label: patients-admissions
con <- dbConnect(SQLite(), "mimic_demo.db")

patients <- read_csv(
  "~/Documents/Spring 26/MEDB-5530/data/mimic-iii-demo/PATIENTS.csv"
)

dbWriteTable(con, "patients", patients, overwrite = TRUE)

admissions <- read_csv(
  "~/Documents/Spring 26/MEDB-5530/data/mimic-iii-demo/ADMISSIONS.csv"
)

dbWriteTable(con, "admissions", admissions, overwrite = TRUE)
glimpse(patients)
glimpse(admissions)
```
## Now we can do a Left join so that we can retain the patient information but add the admission time if available.
```{r}
#| label: left-join
cox <- dbConnect(SQLite(), "mimic_demo.db")
patient_admittime <- dbGetQuery(conn=cox, "
SELECT 
  a.subject_id, 
  p.gender, 
  datetime(a.admittime, 'unixepoch') AS admittime_dt
FROM admissions a
LEFT JOIN patients p
  ON a.subject_id = p.subject_id
LIMIT 10;
")

patient_admittime
dbDisconnect(cox)
```
### Interpretation of the results
We now have a table that shows the subject ID, the patients gender, and the 
time that the patient was admitted for the first 10 patients. 

#### Interpretation of the code
Same as Module 6, we have to convert the unixepoch date into a format that we
as humans can easily identify, that is why we created the admittime_dt to see
exactly what date that patient was admitted. 

# Q2. Get a count of the number of records in the patients table and admissions                       table. How many rows will be retrieved if you do a cross join for patient                                and admissions table.
```{r}
#| label: countsa
cox <- dbConnect(SQLite(), "mimic_demo.db")
counts_a <- dbGetQuery(conn=cox, "
SELECT COUNT(*) AS total_admissions FROM admissions;
")

counts_a
dbDisconnect(cox)
```
```{r}
#| label: countsb
cox <- dbConnect(SQLite(), "mimic_demo.db")
counts_p <- dbGetQuery(conn=cox, "
SELECT COUNT(*) AS total_patients FROM patients;
")

counts_p
dbDisconnect(cox)
```
### Interpretation of output
The admissions data has 129 rows and the patients data has 100 rows. If we were
to do a cross-join, we should have 129*100= 12,900 rows. After checking below,
we can see that we now have 12,900 rows. 

```{r}
#| label: cross-join
cox <- dbConnect(SQLite(), "mimic_demo.db")
crossjoin <- dbGetQuery(conn=cox, "
SELECT COUNT(*) AS total_rows
FROM patients p
CROSS JOIN admissions a;
")

crossjoin
dbDisconnect(cox)
```

# Q3. List all admissions that do not have a diagnosis.                                                  Hint: Use LEFT JOIN + WHERE d.icd9_code IS NULL.
## Checking that diagnosis database is loaded.
```{r}
#| label: checking-diagnosis
diagnosis <- read_csv(
  "~/Documents/Spring 26/MEDB-5530/data/mimic-iii-demo/DIAGNOSES_ICD.csv"
)

dbWriteTable(con, "diagnosis", diagnosis, overwrite = TRUE)
glimpse(diagnosis)
```
## Checking the count of null values first 
```{r}
#| label: checking-null
cox <- dbConnect(SQLite(), "mimic_demo.db")
null <- dbGetQuery(conn=cox, "
SELECT COUNT(*) AS missing_icd9
FROM diagnosis
WHERE icd9_code IS NULL;
")

null
dbDisconnect(cox)
```
## There were no null values, so we should see an empty table here.
```{r}
#| label: null-icd
cox <- dbConnect(SQLite(), "mimic_demo.db")
null_icd <- dbGetQuery(conn=cox, "
SELECT a.subject_id, a.hadm_id, d.icd9_code
FROM admissions a
LEFT JOIN diagnosis d
ON a.subject_id = d.subject_id AND a.hadm_id = d.hadm_id
WHERE d.icd9_code IS NULL;
")

null_icd
dbDisconnect(cox)
```
### Interpretation of output
There were no missing icd_9 codes within this dataset. The first chunk of code
checked to see if there were, but then we used a left join to answer the question
to show that there are no null icd-9 codes therefore we are left with a table 
that has 3 columns (subject ID, hadm ID, and icd-9) but 0 rows because there are
no null values. 