---
title: "todd_5530_m6"
author: "Reagan Todd"
format: 
  html:
    embed-resources: true
execute: 
  error: true
---
```{r}
#| label: setup
#| message: false
#| warning: false
R.version.string
Sys.Date()
library(tidyverse)
library(DBI)
library(RSQLite)
library(readr)
```

### This is a demonstration of how to join tables in SQL created
### 2026-03-27 by Reagan Todd. The copyright is for educational uses only. 

# Q1. Use the labevents table in the MIMICIII to select glucose levels (WHERE itemid = 50809). Use CASE WHEN to categorize glucose levels, if valuenum > 140 categorize as 'High' otherwise categorize as 'Normal_or_Low'.
### First we need to add labevents to our MIMICIII database:
```{r}
#| label: labevents-import
con <- dbConnect(SQLite(), "mimic_demo.db")

labevents <- read_csv(
  "~/Documents/Spring 26/MEDB-5530/data/mimic-iii-demo/LABEVENTS.csv"
)

dbWriteTable(con, "labevents", labevents, overwrite = TRUE)
```
### Now we can categorize the glucose
```{r}
#| label: categorize-glucose
cox <- dbConnect(SQLite(), "mimic_demo.db")
labevents <- dbGetQuery(conn=cox, "
SELECT
  subject_id,
  valuenum AS glucose,
  CASE
    WHEN valuenum > 140 THEN 'High'
    ELSE 'Normal_or_low'
  END AS glucose_category
FROM labevents
WHERE itemid = 50809 -- Glucose
AND valuenum IS NOT NULL
LIMIT 10;")

labevents
dbDisconnect(cox)
```
### Interpretation of output
We can now see that we have the glucose value and a glucose category. 
In just the first 10 rows, we can see that we only have two values under 140
those are categories "Normal_or_low" and we have eight of "High" category. 
Just from the first 10 rows we can see that the case-when conditional formatting
worked. 

#### Interpretation of code
Here we see subject ID, glucose level and the glucose category. We categorized
the glucose into category using the [case where][ref01] where we stated that
if the valuenum (glucose) was >140 then show in category high show in the category of normal or low. A good reference to understanding case-when and
else can be found [here][ref02]. 

[ref01]:https://hightouch.com/sql-dictionary/sql-case
[ref02]:https://www.thoughtspot.com/sql-tutorial/sql-case

# Q2. Use INNER JOIN in MIMIC-III to join patients table and and Admissions table. Get age, gender and admission details for patients. (limit 10 rows)
### First I want to ensure that my data is loaded correctly for patients
### and admissions. 
```{r}
#| label: patients-admissions
con <- dbConnect(SQLite(), "mimic_demo.db")

patients <- read_csv(
  "~/Documents/Spring 26/MEDB-5530/data/mimic-iii-demo/PATIENTS.csv"
)

dbWriteTable(con, "patients", patients, overwrite = TRUE)

admissions <- read_csv(
  "~/Documents/Spring 26/MEDB-5530/data/mimic-iii-demo/ADMISSIONS.csv"
)

dbWriteTable(con, "admissions", admissions, overwrite = TRUE)
```
### Now we can inner join
#### Note this will find the age at admission due to it not being provided.
```{r}
#| label: inner-join
cox <- dbConnect(SQLite(), "mimic_demo.db")
patients_admissions <- dbGetQuery(conn=cox, "
SELECT
  p.subject_id,
  p.gender,
  a.hadm_id,
  a.admittime,
  a.dischtime,
  p.dob,
  DATE(p.dob, 'unixepoch') AS birth_date,
  DATE(a.admittime, 'unixepoch') AS admit_date,
  CAST(strftime('%Y', a.admittime, 'unixepoch') AS INTEGER) AS admit_year,
  ROUND(
    (julianday(a.admittime, 'unixepoch') - julianday(p.dob, 'unixepoch')) / 365.25,
    1
  ) AS age_at_admit
FROM patients p
INNER JOIN admissions a
  ON p.subject_id = a.subject_id
LIMIT 10; ")


patients_admissions 
dbDisconnect(cox)
```
### Interpretation of output
Here we now have an age_at_admit column. This shows the
age at which a person was when they were admitted in years. For the first 10 rows we can see most participants were in their 70's or 80's with a few under 50. The age of 300 represents an individual who was older than 89 years of age at admission. 

#### Interpretation of code
[Date][ref03] is the SQLite version of the TO_DATE as shown in lecture. It returns the date as text format of YYYY-MM-DD.
[CAST][ref04] allowed us to change the format of admit_year from any other format into an integer.  
[julianday][ref03] this is the julian day, the fractional number since a reference day. 
[unixepoch][ref03] returns a unix timestamp which is seen in some of the records where we get the number of seconds since a predetermined date. 
Lastly, the [strftime()][ref03] function for '%Y' returns just the year, since we wanted to calculate age we can use the year to determine that. 

[ref03]:https://sqlite.org/lang_datefunc.html
[ref04]:https://www.dbvis.com/thetable/sql-cast-function-everything-you-need-to-know/

# Q3. Use the MIMICIII schema and write a SQL query that will list all the table names.
```{r}
#| label: table-names
cox <- dbConnect(SQLite(), "mimic_demo.db")
table_names <- dbGetQuery(conn=cox, "
SELECT name AS table_name
FROM sqlite_master
WHERE type='table'
ORDER BY name; ")

table_names
dbDisconnect(cox)
```
### Interpretation of output
As you can see, we have only used the three tables as shown above. This version shows just the tables that I have created in the past 6 modules using the MIMICIII database. 

## Using the mimiciii.db provided, we can see all of the tables in mimiciii. 
```{r}
#| label: entire-db
cox <- dbConnect(SQLite(), "~/Documents/Spring 26/MEDB-5530/data/mimiciii.db")

tables <- dbGetQuery(conn=cox, "
  SELECT name AS table_name
  FROM sqlite_master
  WHERE type='table'
  ORDER BY name;
")

tables
dbDisconnect(cox)
```
# Q4. Use the MIMICIII schema and write a SQL query that will list all the field names in the admissions table.
```{r}
#| label: admissions-fields
cox <- dbConnect(SQLite(), "mimic_demo.db")
table_fields <- dbGetQuery(conn=cox, "
PRAGMA table_info('admissions'); ")

table_fields
dbDisconnect(cox)
```
### Interpretation of output
Now we see the 19 columns as fields from the admissions data. This lists all of the columns. This is very similar to the R version of "colnames()". 

#### Interpretation of code
Here I used [PRAGMA_table_info][ref08] this allows us to return a single row for each column name. This is specific to SQLite and is comparable to the function information_schema.columns shown in lecture. If we just want to see the list of column names, use the syntax below. 

[ref08]:https://sqlite.org/pragma.html#pragma_table_xinfo

```{r}
#| label: admissions-field
cox <- dbConnect(SQLite(), "mimic_demo.db")
table_field <- dbGetQuery(conn=cox, "
SELECT name
FROM pragma_table_info('admissions'); ")

table_field
dbDisconnect(cox)
```