---
title: "todd-5530-m9"
author: "Reagan Todd"
format:
  html:
    embed-resources: true
execute: 
  error: true
---
### This is a demonstration of how to find mismatches in SQL.                                         Created 2026-04-03 by Reagan Todd.                                                                       The copyright is for educational uses only.

```{r}
#| label: setup
#| message: false
#| warning: false
R.version.string
Sys.Date()
library(tidyverse)
library(DBI)
library(RSQLite)
library(readr)
```

# Use the MIMIC III database.
# Q1. Do an inner join of diagnoses_icd and admissions table to find matching diagnoses & admissions. Display the first ten rows of data only.
Use tables
diagnoses_icd  Diagnosis codes per admission (subject_id, hadm_id)
admissions  Admission info (subject_id, hadm_id, admittime, etc.)
use hadm_id to join the tables and detect mismatches.
## First we ensure out database is set to use
```{r}
#| label: checking-database
con <- dbConnect(SQLite(), "mimic_demo.db")

diagnosis <- read_csv(
  "~/Documents/Spring 26/MEDB-5530/data/mimic-iii-demo/DIAGNOSES_ICD.csv"
)

dbWriteTable(con, "diagnosis", diagnosis, overwrite = TRUE)

admissions <- read_csv(
  "~/Documents/Spring 26/MEDB-5530/data/mimic-iii-demo/ADMISSIONS.csv"
)

dbWriteTable(con, "admissions", admissions, overwrite = TRUE)
glimpse(admissions)
glimpse(diagnosis)
```

## Now we can answer the question 
```{r}
#| label: matches
cox <- dbConnect(SQLite(), "mimic_demo.db")
matches <- dbGetQuery(con, "
SELECT 
    d.subject_id,
    d.hadm_id,
    d.icd9_code,
    datetime(a.admittime, 'unixepoch') AS admittime_dt
FROM diagnosis d
INNER JOIN admissions a
    ON d.hadm_id = a.hadm_id;
")

glimpse(matches)
dbDisconnect(cox)
```
#### Interpretation of the code
Same as Module 6, we have to convert the unixepoch date into a format that we
as humans can easily identify, that is why we created the admittime_dt to see
exactly what date that patient was admitted. 

## To find the number of matching records
```{r}
#| label: matches2
cox <- dbConnect(SQLite(), "mimic_demo.db")
matches_count <- dbGetQuery(con, "
SELECT COUNT(*) AS total_matches
FROM diagnosis d
INNER JOIN admissions a
    ON d.hadm_id = a.hadm_id;
")

matches_count
dbDisconnect(cox)
```
### Interpretation of output
We can see the first 10 rows of the data from the first chunk of code, and using 
the second we can see that we were able to obtain 1,761 matching records. Since 
we used an inner join, we can calculate the number of mismatches that the join
omitted:
```{r}
#| label: mismatch_count1
cox <- dbConnect(SQLite(), "mimic_demo.db")
mismatch_count1 <- dbGetQuery(con, "
SELECT COUNT(*) AS total_diagnosis FROM diagnosis
")

mismatch_count1
dbDisconnect(cox)
```
### Interpretation of this output
With a row total of 1,761 for the total diagnosis, it looks like there were no
mismatches between the diagnosis table and admissions table.

# Q2 Write a sql program to select all Diagnoses and Admissions using diagnoses_icd and admissions  table. Hint: Use Full Outer join.
```{r}
#| label: full-outer
cox <- dbConnect(SQLite(), "mimic_demo.db")
full_outer <- dbGetQuery(con, "
SELECT 
    d.*,
    a.* 
FROM diagnosis d
LEFT JOIN admissions a
    ON d.hadm_id = a.hadm_id

UNION

SELECT 
    d.*,
    a.*
FROM admissions a
LEFT JOIN diagnosis d
    ON d.hadm_id = a.hadm_id
")

head(full_outer, 10)
dbDisconnect(cox)

```

#### Interpretation of code
We use the [union][ref01] term here, this term allows us to use two select 
statements which combines those queries and removes duplicates. This is 
primarily useful in R Quarto where a SQL query does not allow for a traditional
full outer join.

[ref01]: https://www.geeksforgeeks.org/sql/sql-union-operator/


### Interpretation of the output
Here we can see that we now have 24 columns (limited to 10 rows) of patient data
using a full outer join to join the admissions and diagnosis data sets. We
will have 1,761 observations or rows and 24 columns. 

# Q3 Which join can be used to detect all match and mismatch cases.
### Answer to question 3
A full outer join as done in question 2 allows us to detect all matches and mis-
matches due to its ability to join all of the data regardless of a match in the
next table. 

# Q4. Which patients(or subject_id) appear in both admissions and diagnoses_icd? Hint use two select statements and the intersect statement.
```{r}
#| label: intersect
cox <- dbConnect(SQLite(), "mimic_demo.db")
intersect <- dbGetQuery(con, "
SELECT DISTINCT subject_id
FROM admissions

INTERSECT

SELECT DISTINCT subject_id
FROM diagnosis;
")
glimpse(intersect)
dbDisconnect(cox)
```
#### Interpretation of the code 
Here we used the [intersect][ref02] term which allows us to return only the 
records that were found within both select statements. This essentially shows
us what is in the center of a vendiagram. 

### Interpretation of the output
We have 100 patients who appear both in the diagnosis and admissions data. 
