---
title: "5508 module04"
author: "Reagan Todd"
format: 
  html:
    embed-resources: true
execute: 
  error: true
---

This program (created 2026-02-10 by Reagan Todd) shows a brief introduction to [null values][ref01] in SQL. This belongs in the public domain so please use however you see fit.

[ref01]: https://www.coginiti.co/tutorials/beginner/sql-is-null/

```{r}
#| label: setup
#| message: false
#| warning: false
R.version.string
Sys.Date()
library(tidyverse)
library(DBI)
library(RSQLite)
library(readr)
```
## creating the MIMIC III database from the demo folder 
### create a new SQLite database file for patients table
```{r}
con <- dbConnect(SQLite(), "mimic_demo.db")

patients <- read_csv(
  "~/Documents/Spring 26/MEDB-5530/data/mimic-iii-demo/PATIENTS.csv"
)

dbWriteTable(con, "patients", patients, overwrite = TRUE)
```

# Q1. Verify that gender in patients table has no missing values
```{r}
#| label: patient_gender
cox <- dbConnect(SQLite(), "mimic_demo.db")
patients <- dbGetQuery(conn=cox, "
    select *
    from patients
    where gender is null;
")
patients
dbDisconnect(cox)
```
### The output of 0 rows confirms that there are no missing values for gender in the patient table.  

# Q2. Count the number of missing values for dod_ssn in patients table
```{r}
#| label: patient_dod_ssn
cox <- dbConnect(SQLite(), "mimic_demo.db")
patients <- dbGetQuery(conn=cox, "
    select *
    from patients
    where dod_ssn is null;
")
patients
dbDisconnect(cox)
```
### There are 23 rows for dod_ssn where there is a null value. This means there are 23 missing or null values for dod_ssn in the patients table. 

## To answer Q3, we need to load in the DRGCODES.csv into the database. 
```{r}
con <- dbConnect(SQLite(), "mimic_demo.db")

drug_codes <- read_csv(
  "~/Documents/Spring 26/MEDB-5530/data/mimic-iii-demo/DRGCODES.csv"
)

dbWriteTable(con, "drug_codes", drug_codes, overwrite = TRUE)
```

# Q3. Find the drg_code of the drgcodes table where drg_severity is null.
```{r}
#| label: drug_severity
cox <- dbConnect(SQLite(), "mimic_demo.db")
drug_codes <- dbGetQuery(conn=cox, "
    select 
      drg_code,
      drg_severity
    from drug_codes
    where drg_severity is null;
")
drug_codes
dbDisconnect(cox)
```
### There are 129 null or missing values for drug severity and listed above are the drug codes (drg_code) for those missing drug severities. 