---
title: "todd-5530-M5"
author: "Reagan Todd"
format: 
  html:
    embed-resources: true
execute: 
  error: true
---
## File details
This program (created 2026-03-06 by Reagan Todd) shows a brief introduction 
to single row functions in SQL. This belongs in the public domain so please 
use however you see fit.

```{r}
#| label: setup
#| message: false
#| warning: false
R.version.string
Sys.Date()
library(tidyverse)
library(DBI)
library(RSQLite)
library(readr)
```
# Q1 Use MIMIC-III Database
## a. Round the value 29.526 to two decimal places.
```{r}
#| label: rounding
cox <- dbConnect(SQLite(), "mimic_demo.db")
rounding <- dbGetQuery(conn=cox, "
    select round(29.526, 2);
    ")
rounding
dbDisconnect(cox)
```
### Interpretation of results
29.526 rounded to 2 decimal places using the round() is 29.53

## Helpful links for tackling rounding
[Round][ref1] function allows for us to round a numeric values within our SQL 
query. It also allows for better readability and control in displaying numbers.

[ref1]: https://hightouch.com/sql-dictionary/sql-round

## b. Truncate the value 29.526 to zero decimal places.
```{r}
#| label: truncating
cox <- dbConnect(SQLite(), "mimic_demo.db")
truncating <- dbGetQuery(conn=cox, "
    select cast(29.526 as INT);
    ")
truncating
dbDisconnect(cox)
```
### Interpretation of results
Truncating 29.526 provides us with an output of 29. This essentally removes the
decimal places. 

## Helpful links to truncation
The SQLite in R does not allow for the use of [TRUNC()][ref2] like it does in 
snowflake so we must use [CAST()][ref3]. Truncation allows for the removal of 
decimal places, whereas cast removes decimals, forces values to 
behave differently (text as number) or changes numeric precision.

[ref2]: https://www.scaler.com/topics/trunc-function-in-sql/
[ref3]: https://www.scaler.com/topics/sql-cast/

## c. Calculate the remainder of the value 2500 after it is divided by 500
```{r}
#| label: remainder
cox <- dbConnect(SQLite(), "mimic_demo.db")
remainder <- dbGetQuery(conn=cox, "
    select (2500 % 500);
    ")
remainder
dbDisconnect(cox)
```
### Interpretation of results
The output of 0 shows that there should be no remainders as 500 can go into
2500 5 times. 

## Helpful link to tackling remainders
SQLite uses % for division whereas other DBMS use the MOD() function. [Both][ref4] allow
for us to divide one number by another and find the remainder. Either function is
not great with handling null values. 

[ref4]: https://www.sherloqdata.io/sql-functions/mod-in-sql

# Q2. Use MIMIC-III Database. Decode Admission Type in admissions Table. Admission_type has values like:
EMERGENCY
URGENT
ELECTIVE
Others
Map them:
EMERGENCY Visit
URGENT Visit
Scheduled Admission
Others
```{r}
#| label: decoding
cox <- dbConnect(SQLite(), "mimic_demo.db")
admission_category <- dbGetQuery(cox, "
select
    admission_type,
    case
        when admission_type = 'EMERGENCY' then 'Emergency Visit'
        when admission_type = 'URGENT' then 'Urgent Visit'
        when admission_type = 'ELECTIVE' then 'Scheduled Admission'
        else 'Others'
    end AS admission_category
from admissions
limit 10;
")
admission_category
dbDisconnect(cox)
```
### Interpretation of results
Using the case function allows us to categorize our admission types into 
admission categories. This results in an if/then/else format to match the cases
found in admission_type to our provided admission_category. 

## Helpful links for case/decode
SQLite uses [case][ref6] where we formulate an when/else structure to get our
categories. Snowflake uses [decode][ref7] where we provide the provided name seperated 
by a comma then our preffered name. 

[ref6]: https://hightouch.com/sql-dictionary/sql-case
[ref7]: https://hightouch.com/sql-dictionary/sql-decode